Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-01 analysis/paper/article.qmd
add table showing breakdown of jobs by type, rank, tenure status

library(tidyverse)
library(here)
library(ggbeeswarm)
# This CSV file was downloaded from our data sheet here
# https://docs.google.com/spreadsheets/d/1Jwe3UqJyedrV-QWlwR_44__t4xBVrCfxGyhXdi3E0sg/edit?resourcekey#gid=1686084773
# note that you may need to download it again to get the latest updates!

@@

We collected data from `r total_number_of_ads_in_our_sample` ads for tenure-track jobs in archaeology posted during 2013-2023. We focus our analysis here on the `r total_number_of_us_ads_in_our_sample` ads for positions at US universities. @fig-show-basic-plots shows the count of ads for each year. Assistant Professor jobs are consistently the most common title and rank of positions advertised, while open rank or full professor are the least frequent. The ratio of tenure-track to non-tenure-track positions is generally well above one; in other words, this data set is dominated by tenure-track positions. Only academic year 2013-2014 had more non-tenure-track positions than tenure-track positions, which was followed by an upward trend peaking at 2018-2019 and then declining again into the present.

@@

>>>>

library(tidyverse)
library(here)
library(ggbeeswarm)
library(flextable)
# This CSV file was downloaded from our data sheet here
# https://docs.google.com/spreadsheets/d/1Jwe3UqJyedrV-QWlwR_44__t4xBVrCfxGyhXdi3E0sg/edit?resourcekey#gid=1686084773
# note that you may need to download it again to get the latest updates!

@@

We collected data from `r total_number_of_ads_in_our_sample` ads for tenure-track jobs in archaeology posted during 2013-2023. We focus our analysis here on the `r total_number_of_us_ads_in_our_sample` ads for positions at US universities. @fig-show-basic-plots shows the count of ads for each year. @tbl-show-basic-counts shows the breakdown by different job types, ranks, and tenure status. Assistant Professor jobs are consistently the most common title and rank of positions advertised, while open rank or full professor are the least frequent. The ratio of tenure-track to non-tenure-track positions is generally well above one; in other words, this data set is dominated by tenure-track positions. Only academic year 2013-2014 had more non-tenure-track positions than tenure-track positions, which was followed by an upward trend peaking at 2018-2019 and then declining again into the present.

@@

```{r}
#| label: tbl-show-basic-counts
#| tbl-cap: "Breakdown of counts of job ads by rank and tenure status. Values in parentheses are for the United States only. TT = tenure-track, NTT = non-tenure-track. Not all ads include unambiguous information about tenure status, so the sum of TT and NTT does not equal the sum of all jobs in all ranks."

#---------------------------------------------
# want a table to show values for fig 1

prop_by_job_title_per_year_tbl_wide <-
prop_by_job_title_per_year_tbl %>%
  select(-prop) %>%
  pivot_wider(names_from = job_title_simple,
              values_from = n)

prop_by_job_title_per_year_us_only_tbl <-
jobdata_us_only %>%
  group_by(year_ad_posted) %>%
  count(job_title_simple) %>%
  mutate(prop = n / sum(n)) %>%
  select(-prop) %>%
  pivot_wider(names_from = job_title_simple,
              values_from = n)

# combine tables with total (US) numbers

# 1. Join the two tables by year, adding suffixes to distinguish columns
combined_data <- left_join(
  prop_by_job_title_per_year_tbl_wide,
  prop_by_job_title_per_year_us_only_tbl,
  by = "year_ad_posted",
  suffix = c("_wide", "_us") # Suffixes for columns from each table
)

# 2. Identify the original value column names (excluding the join key)
value_cols <- setdiff(names(prop_by_job_title_per_year_tbl_wide), "year_ad_posted")

# 3. Mutate to create the combined columns
#    Iterate through the original value column names
for (col_name in value_cols) {
  # Construct the names of the columns from the joined table
  col_wide_name <- paste0(col_name, "_wide")
  col_us_name <- paste0(col_name, "_us")

  # Use mutate to create a new column (overwriting the original name)
  # with the combined string "Value_Wide (Value_US)"
  # Handle NA values by replacing them with a placeholder like "-"
  combined_data <- combined_data %>%
    mutate(
      !!col_name := paste0(
        # Coalesce NA to "-" before pasting
        coalesce(as.character(!!sym(col_wide_name)), "-"),
        " (",
        coalesce(as.character(!!sym(col_us_name)), "-"),
        ")"
      )
    )
}

# 4. Select only the year column and the newly created combined columns
final_combined_table <- combined_data %>%
  select(year_ad_posted, all_of(value_cols))

# let's add TT and NTT columns to that
count_of_tt_jobs_per_year_from_our_form_tbl <- 
count_of_tt_jobs_per_year_from_our_form %>%
  select(year_ad_posted = year,
         TT = n_tt_jobs,
         NTT = n_non_tt_jobs) %>%
  left_join(final_combined_table) %>%
  relocate(TT, NTT, .after = last_col()) %>%
  relocate(`Full Professor`, .after = `Associate Professor`) %>%
  rename( `Year ad posted` = year_ad_posted) %>%
  # get numeric year, sort, then remove it
  mutate(year_num = parse_number(str_sub(`Year ad posted`, 6L))) %>%
  arrange(year_num) %>%
  select(-year_num)

ft <- flextable(count_of_tt_jobs_per_year_from_our_form_tbl)

# Apply formatting for a clean, professional look similar to booktabs
ft <- ft %>%
  # Apply a theme with minimal lines, good header/footer separation
  theme_booktabs() %>% 
  # Adjust column widths to fit
  autofit() %>% fit_to_width(7) %>% 
  # Left-align all columns (adjust if needed)
  align(align = "left", part = "all") %>% 
  # Center numeric columns (adjust selector)
  align(j = ~ . -year_ad_posted, align = "center", part = "all") %>% 
  # Add some vertical padding
  padding(padding.top = 4, padding.bottom = 4, part = "all")

# Print the flextable object - it will render as a table in Word
ft
```
